Exploring probability distributions for bivariate temporal granularities

### 2019-12-15 main.Rmd
choice of plots section update

## Measured variable

@@

Drawing from the grammar of graphics, if <$A_i$, $B_j(s)$, $v(s)$> $i \in {1, 2, \dots, n}$ and $j \in {1, 2, \dots, m}$ serves as the basis of visualizing the probability distribution of the measured variable, the following components of the layered grammar can be specified:

@@

## Choice of statistical transformations and geometric objects

@@

There are several ways to summarize a statistical distributions. The entire distribution can be estimated through density estimates using non-paramteric density estimation methods or only summarized using descriptive statistics. The latter do not allow us to see the shape, skewness, nature of the tail or multi-modality with so much clarity as the former. However, they avoid much clutter, where some specific probabilities representing typical and extreme behavior are on focus. Non-parametric density based plots can show the entire distribution, unlike discretizing and showing only parts of it. These displays depend on the specific estimation procedure (e.g., kernel density estimate) as well as on additional smoothing parameters and hence can vary for the same data set. In practice, the densities are estimated reasonably with at least 30 observations. However, even with large sample size, choosing too large a value for bandwidth can cause the estimate to oversmooth the data. The summaries based on descriptive statistics remain free of any tuning parameters or Kernels.

@@

@@

### Descriptive statistics based methods

Most commonly used techniques include the box plots [@Tukey1977-jx] which convey statistical features such as the median, quartile boundaries, hinges, whiskers and extreme outliers. These displays are based on descriptive statistics, for example, $([n/2] + 1)/2^{th}$ and $(n + 1 - ([n/2] + 1)/2)^{th}$ order statistics are taken as estimates of the quartiles. Due to wide spread popularity and simplicity in implementation, a number of variations are proposed to the original one which provides alternate definitions of quantiles, whiskers, fences and outliers. Notched box plots [@Mcgill1978-hg] has box widths proportional to the number of points in the group and display confidence interval around medians aims to overcome some drawbacks of box plots.<!-- Challenge of boxplot --> The standard box plot and all of it's variations are helpful to get an idea of the distribution at a glance. However, for data less than 1000 observations, detailed estimates of tail behavior beyond the quartiles are not trustworthy. Also, the number of outliers is large for larger data set since number of outliers is proportional to the number of observations.

@@

### Non-parametric density based methods

Traditional ways to visualize densities include violin plots [@Hintze1998-zi]. The shape of the violin represents the density estimate of the variable. The more data points in a specific range, the larger the violin is for that range. Adding two density plots gives a symmetric plot which makes it easier to see the magnitude of the density and compare across categories, enabling easier detection of clusters or bumps within a distribution.

>>>>

## Measured variable {#sec:measured-variable}

@@

Drawing from the grammar of graphics, if <$A_i$, $B_j(s)$, $v(s)$> $i \in {1, 2, \dots, n}$ and $j \in {1, 2, \dots, m}$ serves as the basis of visualizing the distribution of the measured variable, the following components of the layered grammar can be specified:

@@

## Choice of plots

@@

Choice of plots are dictated by the statistical transformations and geometric objects used for the visualization. The basic plot choice is display of density. In Section \ref{sec:measured-variable}, it is dicussed that these densities could be estimated using Kernel density estimates or some descriptive statistics. The latter do not allow us to see the shape, skewness, nature of the tail or multi-modality with so much clarity as the former. However, they avoid much clutter and help us to focus on some specific properties of the data. as well as on additional smoothing parameters and hence can vary for the same data set. In practice, the densities are estimated reasonably with at least 30 observations. However, even with large sample size, choosing too large a value for bandwidth can cause the estimate to oversmooth the data. The summaries based on descriptive statistics remain free of any tuning parameters or Kernels, which is sometimes more desirable for exploration.

@@

<!-- Non-parametric density based plots can show the entire distribution, unlike discretizing and showing only parts of it. These displays depend on the specific estimation procedure (e.g., kernel density estimate)  -->

@@

### Descriptive statistics based displays

Most commonly used displays include the box plots [@Tukey1977-jx] which convey statistical features such as the median, quartile boundaries, hinges, whiskers and extreme outliers. These displays are based on descriptive statistics, for example, $([n/2] + 1)/2^{th}$ and $(n + 1 - ([n/2] + 1)/2)^{th}$ order statistics are taken as estimates of the quartiles. Due to wide spread popularity and simplicity in implementation, a number of variations are proposed to the original one which provides alternate definitions of quantiles, whiskers, fences and outliers. Notched box plots [@Mcgill1978-hg] has box widths proportional to the number of points in the group and display confidence interval around medians aims to overcome some drawbacks of box plots.<!-- Challenge of boxplot --> The standard box plot and all of it's variations are helpful to get an idea of the distribution at a glance. However, for data less than 1000 observations, detailed estimates of tail behavior beyond the quartiles are not trustworthy. Also, the number of outliers is large for larger data set since number of outliers is proportional to the number of observations.

@@

### Kernel density based displays

Traditional ways to visualize kernel densities include violin plots [@Hintze1998-zi]. The shape of the violin represents the density estimate of the variable. The more data points in a specific range, the larger the violin is for that range. Adding two density plots gives a symmetric plot which makes it easier to see the magnitude of the density and compare across categories, enabling easier detection of clusters or bumps within a distribution.
